Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-02 application.tex
updating

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $r_\mathrm{AU}\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
  \beta(z; \khat) =
  &\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)r_\mathrm{AU}^{-57/14}\notag\\
&\times\left[\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}r_\mathrm{AU}^{33/7}+\frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right].     
\end{align}
We refer to the case with the scalings
$\hat{\Sigma}=\hat{\kappa}_d=\hat{T}=1$ as the Minimum Mass Solar
Nebulae (MMSN). The physical thermal relaxation timescale
$\beta$ is height-dependent, whilst for analytical discussion we
assumed it is constant. Nevertheless, it is instructive to evaluate
and compare with the criterion Eq. \ref{iso_vsi_cond} for the fiducial  
disk model,  
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is necessary for the
fundamental VSI when $\khat\gg 1$. %For $\khat\to0$, this
%becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
%\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn}  with $\mu
=2.33$ and $\gamma=1.4$.  We set the scaling $\hat{T}=\hat{\Sigma}=1$
and consider three opacity scales: $\hat{\kappa}_d=0.1, \,1 $ and $10$
at two heights: $z=0$ and $z=3H$. We also plot Eq. \ref{bcrit_mmsn} in
each panel for comparison.

Consider first the MMSN shown in the middle panels of
Fig. \ref{mmsn_bcrit_bcool}. The critical thermal timescale,
Eq. \ref{bcrit_mmsn}, as derived from our analytical discussion, is
most useful to assess the VSI in the optically thin regime (where the
curves overlap), since in this limit $\beta$ does not depend on
height. We can therefore conclude that the VSI can operate at tens of
AU in the MMSN.

In the inner few AU of the MMSN, however, $\beta$  has a
strong dependence on height because the disk is optically thick. Given
a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at the
midplane, but eventually $\beta<\beta_\mathrm{crit}$ with increasing
height. Our analytical criteria is not valid for variable $\beta$, but
given that the VSI depends on vertical shear, an effect that becomes
important away from the disk midplane, the inner disk may still be
able to develop the VSI, but it is likely restricted to the disk
surface.

Now consider lowering the opacity, as shown in the top panels of
Fig. \ref{mmsn_bcrit_bcool}. The curves shift inward, but raises also
them. In fact, lowering the opacity by an order of magnitude renders
$\beta > \beta_\mathrm{crit}$ even away from the midplane. This
suggests that fiducial  disk stable to the fundametal mode for the
range of wavenumbers considered.

Raising the opacity, as shown in the lower panels of
Fig. \ref{mmsn_bcrit_bcool}, decreases the thermal timescales, but
also shifts the curves outward, so the fundamental mode is restricted
to larger radii.

%We remark that it is possible to satisfy $\beta<\beta_\mathrm{crit}$
%at small radii by choosing a sufficiently large $\khat$. However, our
%linear calculations suggest growth rates eventually decay as 
%$\khat\to\infty$. Thus, very small-scale VSI in the inner disk is
%probably unimportant.

@@

We remark that it is possible to satisfy $\beta <
\beta_\mathrm{crit}$ by choosing a sufficiently large $\khat$, or 
equivalently a sufficiently small perturbation lengthscale. We
illustrate this by plotting at each radius the value of $\khat$ such
that $\beta = \beta_\mathrm{crit}$. In the inner few AU of the MMSN,
one needs to consider large values of $\khat$ and/or regions away from
the disk midplane.

@@

\subsection{Numerical examples}\label{mmsn_example}
Here we solve the linear stability problem with the
thermal relaxation function defined by Eq. \ref{beta_mmsn}. We
consider the MMSN at $r=50$AU and $r=2$AU.

We first compare $\beta$ at these radii in Fig. \ref{beta_compare}. In
the optically thin outer disk, $\beta$ is independent of $z$ and
$\beta<\beta_\mathrm{crit}$ throughout the disk column. We thus expect
the VSI to operate at $r=50$AU. In the inner disk, $\beta$ depends on 
height such that $\beta\lesssim\beta_\mathrm{crit}$ only for 
$|z|\gtrsim 2H$, which suggest that the VSI will be restricted
restricted to the upper and lower layers in the inner disk.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/beta_compare} 
  \caption{Thermal relaxation timescales in the MMSN at $r=50$AU
    (thick solid) and $r=2$AU (thick dotted) for perturbation
    wavenumbers $\khat=10,\,100$, respectively. The corresponding
    horizontal lines are the critical thermal relaxation timescales
    derived in linear theory. 
    \label{beta_compare}}  
\end{figure}

In Fig. \ref{eigen_compare_mmsn} we compute growth rates of the
fundamental mode at $r=50$AU and $r=2$AU. At $r=50$AU growth rates are
insensitive to $\khat$ for $1\lesssim \khat\lesssim 60$, and
corresponds to growth time $\simeq 40$ local orbits. In the inner disk
at $r=2$AU, we only found unstable modes for $\khat\gtrsim50$. To
obtain a similar growth timescale at $r=2$AU requires $\khat\sim
100$. This is consistent with the expectation that, as we proceed into
the inner disk, it bcomes optically thick, so one needs to consider
smaller perturbation wavelengths to obtain a sufficiently short
thermal timescale.

% In both cases we find that increasing $\khat$ increases the
% perturbation amplitudes near the disk surface. Such modes may be
% affected by boundary conditions so their robustness is unclear.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigen_compare_grow_mmsn} 
  \caption{Growth rates of the fundamental VSI mode in the MMSN at
    $r=50$AU (solid) and $r=2$AU (dotted). 
    \label{eigen_compare_mmsn}}  
\end{figure}

Finally, we compare eigenfunctions for the vertical velocity at these
radii in Fig. \ref{mmsn_eigenvz}. We compare cases with similar growth
rates. At $r=50$AU we have vertical motion throughout the disk column
because the thermal timescale is less than critical. By contrast, at
$r=2$AU there is little vertical motion in $|z|\lesssim 2H$. 
Considering this result with Fig. \ref{beta_compare} suggests, but does
not prove, that the critical thermal timescale gives an indication of
where the VSI operates even when $\beta$ depends on $z$.

%can always go to sufficiently large heights to get beta < beta crit,
%but getting close to boundaries/lower densities

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_50AU}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_2AU}
  \caption{Comparison of vertical velocity perturbations of 
    fundamental VSI modes of similar growth rates in the MMSN at
    $r=50$AU (top) and $r=2$AU (bottom). The real (imaginary) part of
    $\delta v_z$ is shown as 
    the solid (dotted) curves. The eigenfunction has been normalized
    such that $\imag(\delta v_z)=0$ at the vertical boundaries, and
    $\mathrm{max}|\delta v_z|=1$. \label{mmsn_eigenvz}}      
\end{figure}
% only find one unstable mode -> close to marginal 
% larger kx

\subsection{VSI in the outer parts of the MMSN}
To obtain an overall picture of the relevance of the VSI to the MMSN,
we calculate characteristic growth times $t_\mathrm{grow} = 1/\nu$ of 
the fundamental mode in units of the local Keplerian orbital period
$P_\mathrm{orb}= 2\pi/\Omega_k$. The results are shown in 
Fig. \ref{mmsn_overall} for perturbation wavenumbers $5\lesssim
\khat\lesssim 50$. We consider this range because for  
$\khat=O(1)$ or $\khat=O(10^2)$ we found eigenfunctions have large
amplitudes near the vertical boundaries, so such modes may be affected
by boundary conditions.

In the MMSN the fundamental VSI operates at tens of AU, with
characteristic growth times of tens of orbits. %In our example, the VSI
% is most favored at $r\simeq 20\mathrm{AU}$ with a growth time
% $t_\mathrm{grow}\simeq 30$ orbits.
Fig. \ref{mmsn_overall} indicates that with decreasing radius, the VSI
shifts to smaller scales (larger $\khat$), since the disk becomes optically
thick so only small-scale disturbances have sufficiently short thermal
relaxation timescales. At larger radii ($r\gtrsim 
40\mathrm{AU}$), the disk is optically thin so growth times do not depend strongly on   
$\khat$, and the VSI can operate on larger scales (smaller $\khat$).  
However, approaching 
$r\simeq 100\mathrm{AU}$ growth times become long at $O(10^2)$ local orbits,
so the VSI may not be dynamically important in the outermost parts of
the MMSN.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigen_compare_grow.ps}
  \caption{Characteristic growth times of the fundamental VSI mode in
    the MMSN for a range of perturbation wavenumbers.  
    \label{mmsn_overall}}    
\end{figure}

 % \subsection{Vertically-dependent thermal relaxation} 
% Note that for a vertically isothermal disk and opacity $\kappa_d(T)$, 
% we can restore the $z$ dependence in $\beta$ (due to the $\rho$
% dependence in  $\eta$, see Eq. \ref{eta_def}) by making the transformation
% $\beta(\khat)\to\beta(\khat/\sqrt{2\pi}\hat{\rho})\equiv 
% \tilde{\beta}(\khat,z)$ in  Eq. \ref{real_beta}, where
% $\hat{\rho}=\rho/\rho(z=0)$.  We have repeated the example
% calculation in \S\ref{mmsn_example} with $\beta\to\tilde{\beta}$, but
% found similar results.

%We find growth rates
%were reduced, but at most by $\sim 15\%$ at $\khat=1$. Eigenfunctions
%were similar.

%at most by 15%

>>>>

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $r_\mathrm{AU}\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
  \beta(z; \khat) =
  &\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)r_\mathrm{AU}^{-57/14}\notag\\
&\times\left[\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}r_\mathrm{AU}^{33/7}+\frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right].       
\end{align} 
We refer to the case with $\gamma=1.4$, $\mu=2.33$ and the scalings
$\hat{\Sigma}=\hat{\kappa}_d=\hat{T}=1$ as the Minimum Mass Solar
Nebulae (MMSN) and $\beta=\beta_\mathrm{MMSN}$, 
\begin{align}\label{beta_mmsn_simp}
  \beta_\mathrm{MMSN} = 3.9\times10^{-3} r_\mathrm{AU}^{9/14}\left( 1 +
    1.9\times10^7\hat{\rho}^2r_\mathrm{AU}^{-33/7}\khat^{-2}\right). 
\end{align}
% where
% \begin{align}
%   \
% %  \frac{3.7\times10^{-3}}{\mu\left(\gamma-1\right)\hat{\kappa}_d\hat{T}^5}, 
% \end{align}
% and
% \begin{align}
%   \khat_0 = 4.4\times10^{3}\hat{\kappa}_d\hat{T}^2\hat{\Sigma}.
% \end{align}

% The physical thermal relaxation timescale $\beta$ is height-dependent,
% whilst for analytical discussion we assumed it is
% constant. Nevertheless,
It is instructive to compare the above thermal timescales
with the criterion Eq. \ref{iso_vsi_cond} evaluated for the fiducial disk model,    
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is necessary for the
fundamental VSI when $\khat\gg 1$. %For $\khat\to0$, this
% becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
%\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn}  with $\mu
=2.33$ and $\gamma=1.4$. We set the scaling $\hat{T}=\hat{\Sigma}=1$
and consider three opacity scales: $\hat{\kappa}_d=0.1, \,1 $ and $10$
at two heights: $z=0$ and $z=3H$. We also plot Eq. \ref{bcrit_mmsn} in
each panel.

Consider first the MMSN shown in the middle panels of
Fig. \ref{mmsn_bcrit_bcool}. The critical thermal timescale,
Eq. \ref{bcrit_mmsn}, as derived from our analytical discussion, is
most useful to assess the VSI in the optically thin regime (where the
curves overlap), since in this limit $\beta$ does not depend on
height. We thus infer that the VSI can operate at tens of AU in the
MMSN.

In the inner few AU of the MMSN, however, $\beta$ depends on $z$. 
Given a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at
the optically thick midplane, but with increasing height the disk
becomes optically thin and $\beta<\beta_\mathrm{crit}$. Our analytical
criteria is not valid for variable $\beta$, and we must turn to
numerical solutions of the linear to determine whether or not the
fundamental mode operates. We do this in the next
section.

Now consider lowering the opacity, as shown in the top panels of
Fig. \ref{mmsn_bcrit_bcool}. The curves shift inward, but raises also
them. In fact, lowering the opacity by an order of magnitude renders
$\beta > \beta_\mathrm{crit}$ even away from the midplane. This makes
disk stable to the fundametal mode. Raising the opacity, as shown in
the lower panels of Fig. \ref{mmsn_bcrit_bcool}, decreases the thermal
timescales, but shifts the curves outward, so the fundamental mode is
restricted to larger radii.

%but
%given that the VSI depends on vertical shear, an effect that becomes
%important away from the disk midplane, the inner disk may still be
%able to develop the VSI, but it is likely restricted to the disk
%surface.   
%We remark that it is possible to satisfy $\beta<\beta_\mathrm{crit}$
%at small radii by choosing a sufficiently large $\khat$. However, our
%linear calculations suggest growth rates eventually decay as 
%$\khat\to\infty$. Thus, very small-scale VSI in the inner disk is
%probably unimportant.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
by choosing a sufficiently large $\khat$. We
illustrate this by plotting at each radius the value of $\khat$ such
that $\beta = \beta_\mathrm{crit}$. We thus expect that the VSI,
should it occur in the inner disk, would only occur on very small
scales. It is not clear if such modes are robust (see below).

@@

\subsection{VSI in the MMSN}
We have solved the linear problem in the MMSN with the scale and
height dependent thermal relaxation function defined by
Eq. \ref{beta_mmsn_simp}.  We use a vertical
domain $z_\mathrm{max}=5H$. In Fig. \ref{mmsn_overall} we plot growth
timescales $t_\mathrm{grow} \equiv 1/\nu$ associated with the
fundamental mode in units of the local orbital time
$P_\mathrm{orb}\equiv 2\pi/\Omega_k$.

In fact, for $1\lesssim \khat \lesssim 30$ we either only find the
fundamental mode, or it is the most unstable. The associated 
growth rates are also insensitive to $z_\mathrm{max}$. However, for
large $\khat$ and/or $z_\mathrm{max}$, e.g. $\khat=50$ or 
$z_\mathrm{max}=10H$, we also found modes with shorter growth times 
than those shown in Fig. \ref{mmsn_overall}, but these disturbances are 
concentrated near the vertical boundaries
\citep{nelson13,mcnally14}. Such modes may not be robust because    
they are affected by boundary conditions. Furthermore, modes with   
large $\khat$ are likely subject to viscous decay in a real disk.

We therefore take Fig. \ref{mmsn_overall} as a representation of the
VSI in the standard MMSN. We see that growth timescales are $\sim
30$---40 orbits, and the VSI operates from a few AU to tens of
AU. Importantly, it is necessary to consider smaller scales as $r$
decreases as the disk generally becomes optically thick.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigen_compare_grow.ps}
  \caption{Characteristic growth times of the fundamental VSI mode in
    the MMSN.  
    \label{mmsn_overall}}    
\end{figure}

\subsubsection{Comparing the VSI in the inner and outer disk} 
Here we compare the VSI at $r=50$AU and $r=5$AU. We first show in
Fig. \ref{beta_compare}  the thermal timescale $\beta$ for a
perturbation wavenumber $\khat=30$. For comparison we also plot
$\beta_\mathrm{crit}$ at these radii.

 \begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/beta_compare}
  \caption{Thermal relaxation timescales in the MMSN at $r=50$AU
    (thick solid) and $r=5$AU (thick dotted) for $\khat=30$. The
    corresponding thin horizontal lines are the critical thermal
    relaxation timescales derived in linear theory. 
    \label{beta_compare}}
\end{figure}

In the outer disk ($r=50$AU) the disk is optically thin, $\beta$ 
does not depend on height and $\beta <
\beta_\mathrm{crit}$. Accordingly we expect, and find, the VSI to
operate at $r=50$AU. In the inner disk ($r=5$AU) the disk is optically
thick near the midplane and becomes optically thin away from it, such
that $\beta < \beta_\mathrm{crit}$ only for $|z|\gtrsim1.25H$. In
fact, in this case  we also find the VSI to operate, but with
significantly reduced growth rate (as reflected in
Fig. \ref{mmsn_overall}).

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column. This is probably due
to the vertically global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests, but does not prove, that the VSI can operate
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the growth rates are much
smaller than the case with $\beta < \beta_\mathrm{crit}$ for all $z$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_50AU}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_5AU}
  \caption{Vertical velocity perturbations of the VSI in the MMSN at 
    $r=50$AU (top) and $r=5$AU (bottom). The real (imaginary) part of 
    $\delta v_z$ is shown as
    the solid (dotted) curves. The eigenfunction has been normalized
    such that $\imag(\delta v_z)=0$ at the vertical boundaries, and
    $\mathrm{max}|\delta v_z|=1$. \label{mmsn_eigenvz}}
\end{figure}
